~\Large Stickle!

### 2023-05-08 manuscript/paper.Rmd
Update Manuscript
Added some more description to our model specification.

@@

<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon May  1 11:20:28 2023 -->
\begin{table}[ht]
\centering
\begin{tabular}{rl}
  \hline
time & count \\ 
  \hline
  1 & 43 \\ 
    2 & 41 \\ 
    3 & 51 \\ 
    4 & 41 \\ 
    5 & 46 \\ 
    6 & 48 \\ 
    7 & 67 \\ 
    8 & 55 \\ 
    9 & 42 \\ 
   10 & 33 \\ 
   11 & 37 \\ 
   12 & 22 \\ 
   13 & 41 \\ 
   14 & 43 \\ 
   15 & 46 \\ 
   16 & 47 \\ 
   17 & 56 \\ 
   18 & 55 \\ 
   \hline
\end{tabular}
<!-- \label{tab1} -->
\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
\end{table}

@@

# Models  {#sec:models}
## Imputation model 
Let $W$ be sex and $X$ are covariates.  This data set up here is $n_{fossil} + n_{modern}$.  
$W = (W_{obs},W_{mis})$ where $W_{obs}$ and $W_{mis}$ are the observed and missing parts of the covariate sex.  In this setting, $W_{obs}$ perfectly corresponds to the sex of the modern specimens, and $W_{mis}$ perfectly corresponds to the fossil data.  Missing values of sex are imputed by drawing from the posterior predictive distribution $P(W_{mis}|W_{obs}, X)$. Multiple imputation was implemented here using MICE (CITE) with predictive mean matching.  The data were imputed M = 100 times.

## OU model 
After imputing sex, we fit an OU model for different phenotypes of interest. We only look at $n_{fossil}$ fossil observations.  
WLOG we choose one of the columns of $X$ to be the target ophenotype and our analysis is repeated for all phenotypes of interest.

Let $X_{gti}$ be a phenotype of interest for the $i$-th observation $i = 1\ldots,n_{gt}$, in sex $g = 1, 2$ and $t = 1, \ldots, 18$.

$$
X_{gti} = \theta_g + u_{gt} + \epsilon_{gti}
$$
$$
u_{gt} = \kappa u_{g(t-1)} + \nu_{gt}
$$
$$
u_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right)
$$

$$
\nu_{gt} \sim \mathcal{N}(0,\tau^2)
$$

$$
\epsilon_{gti} \sim \mathcal{N}(0,\sigma^2)
$$

@@

$$
\kappa \sim \mathcal{N}(0,25)I_{\{-1 < \kappa < 1\}} \\
$$
$$
\theta_g \sim \mathcal{N}(54,625)I_{\{\theta_g > 0\}} \\
$$

>>>>

\newcommand{\iid}{\overset{iid}{\sim}}

@@

<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon May  1 11:20:28 2023 -->
\begin{table}[ht]
\centering
\begin{tabular}{rl}
  \hline
time & count \\ 
  \hline
  1 & 43 \\ 
    2 & 41 \\ 
    3 & 51 \\ 
    4 & 41 \\ 
    5 & 46 \\ 
    6 & 48 \\ 
    7 & 67 \\ 
    8 & 55 \\ 
    9 & 42 \\ 
   10 & 33 \\ 
   11 & 37 \\ 
   12 & 22 \\ 
   13 & 41 \\ 
   14 & 43 \\ 
   15 & 46 \\ 
   16 & 47 \\ 
   17 & 56 \\ 
   18 & 55 \\ 
   \hline
\end{tabular}
\label{tab1}
\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
\end{table}

@@

# Models  {#sec:models}
## Imputation model 
Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times p$ matrix of the $p$ phenotypes of interest of the fish.  Because $\boldsymbol{W}$ contains missing data, we further define $\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T$ where $\boldsymbol{W}_{obs}$ and $\boldsymbol{W}_{mis}$ are the observed and missing data of the covariate gender, respectively.  In this setting, $\boldsymbol{W}_{obs}$ is a $n_{extant} \times 1$ vector and $\boldsymbol{W}_{obs}$ is a $n_{fossil} \times 1$ vector, perfectly corresponding to the observable gender of the modern extants and the unobservable gender of the fossil data, respectively.

We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{mis}|\boldsymbol{W}_{obs}, \boldsymbol{X})$ using the multiple imputation with chained equations (MICE) algorithm (@MICE) with predictive mean matching.  The imputation algorithm is run to obtain a total of $M = 100$ completed datasets.

## OU model 
After imputing sex, we fit an OU model for different phenotypes of interest. We only look at $n_{fossil}$ fossil observations.  
WLOG we choose one of the columns of $X$ to be the target phenotype and our analysis is repeated for all phenotypes of interest.

Let $W_{ti}$ and $X_{ti}$ be the gender and phenotype of interest for the $i$-th observation in year $t$ for $t = 1, \ldots, 18$ and  $i = 1\ldots,n_{t}$.  We assume that

$$
\begin{aligned}
X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\mu_{ft},\sigma^2) & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{mt},\sigma^2) & W_{ti} = \text{Male} \end{array}\right.
\end{aligned}
$$
We then fit the time and gender dependent means, $\mu_{ft}$ and $\mu_{mt}$, to an autoregressive process of lag 1, which is the discretization of an OU process.  For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for the mean structure defined as is

$$
\begin{aligned}
\mu_{gt} \iid \mathcal{N}(\kappa \mu_{g,t-1} + (1 - \kappa)\theta_g, \tau^2).
\end{aligned}
$$
In this setting, $\theta_g, g \in \{f,m\}$ is the defined as the overall long-run mean of the phenotype of interest for each gender, and $\kappa$ represents how quickly over time $\mu_{gt}$ reverts back to $\theta_g$; the larger the value of $\kappa$, the slower $\mu_{gt}$ reverts back to $\theta_g$.

We also set

$$
\begin{aligned}
\mu_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right).
\end{aligned}
$$
This model choice is to preserve stationarity, i.e. $p(\mu_{gs}) = p(\mu_{gt})$ for $s,t \in \{1,\dots,T\}$.

@@

$$
\kappa \sim \mathcal{N}(0,25)I_{\{0 < \kappa < 1\}} \\
$$
$$
\theta_g \sim \mathcal{N}(54,625)I_{\{\theta_g > 0\}} \\
$$

### 2023-05-08 manuscript/paper.tex
Add Figures
Upload figures discussed in the Cinco de Mayo meeting.
Update Manuscript
Added some more description to our model specification.

\begin{table}[ht]
\centering
\begin{tabular}{rl}
  \hline
time & count \\ 
  \hline
  1 & 43 \\ 
    2 & 41 \\ 
    3 & 51 \\ 
    4 & 41 \\ 
    5 & 46 \\ 
    6 & 48 \\ 
    7 & 67 \\ 
    8 & 55 \\ 
    9 & 42 \\ 
   10 & 33 \\ 
   11 & 37 \\ 
   12 & 22 \\ 
   13 & 41 \\ 
   14 & 43 \\ 
   15 & 46 \\ 
   16 & 47 \\ 
   17 & 56 \\ 
   18 & 55 \\ 
   \hline
\end{tabular}
<!-- \label{tab1} -->
\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
\end{table}

@@

Let \(W\) be sex and \(X\) are covariates. This data set up here is
\(n_{fossil} + n_{modern}\).\\
\(W = (W_{obs},W_{mis})\) where \(W_{obs}\) and \(W_{mis}\) are the
observed and missing parts of the covariate sex. In this setting,
\(W_{obs}\) perfectly corresponds to the sex of the modern specimens,
and \(W_{mis}\) perfectly corresponds to the fossil data. Missing values
of sex are imputed by drawing from the posterior predictive distribution
\(P(W_{mis}|W_{obs}, X)\). Multiple imputation was implemented here
using MICE (CITE) with predictive mean matching. The data were imputed M
= 100 times.

@@

After imputing sex, we fit an OU model for different phenotypes of
interest. We only look at \(n_{fossil}\) fossil observations.\\
WLOG we choose one of the columns of \(X\) to be the target ophenotype
and our analysis is repeated for all phenotypes of interest.

Let \(X_{gti}\) be a phenotype of interest for the \(i\)-th observation
\(i = 1\ldots,n_{gt}\), in sex \(g = 1, 2\) and \(t = 1, \ldots, 18\).

\[
X_{gti} = \theta_g + u_{gt} + \epsilon_{gti}
\] \[
u_{gt} = \kappa u_{g(t-1)} + \nu_{gt}
\] \[
u_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right)
\]

\[
\nu_{gt} \sim \mathcal{N}(0,\tau^2)
\]

\[
\epsilon_{gti} \sim \mathcal{N}(0,\sigma^2)
\]

Priors: \[
\sigma \sim \mathcal{N}(0,2)I_{\sigma > 0}
\] \[
\tau \sim \mathcal{N}(0,2)I_{\tau > 0}
\]

\[
\kappa \sim N(0.5,1)\\
\] \[
\theta_g \sim N(54,400) \\
\]

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Cornault2022}{}}%
Cornuault, Josselin. 2022. {``{Bayesian Analyses of Comparative Data
with the Ornstein--Uhlenbeck Model: Potential Pitfalls}.''}
\emph{Systematic Biology} 71 (6): 1524--40.
\url{https://doi.org/10.1093/sysbio/syac036}.

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{rl}
  \hline
time & count \\ 
  \hline
  1 & 43 \\ 
    2 & 41 \\ 
    3 & 51 \\ 
    4 & 41 \\ 
    5 & 46 \\ 
    6 & 48 \\ 
    7 & 67 \\ 
    8 & 55 \\ 
    9 & 42 \\ 
   10 & 33 \\ 
   11 & 37 \\ 
   12 & 22 \\ 
   13 & 41 \\ 
   14 & 43 \\ 
   15 & 46 \\ 
   16 & 47 \\ 
   17 & 56 \\ 
   18 & 55 \\ 
   \hline
\end{tabular}
\label{tab1}
\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
\end{table}

@@

Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
vector of the covariate gender of the stickleback fish and
\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times p\) matrix
of the \(p\) phenotypes of interest of the fish. Because
\(\boldsymbol{W}\) contains missing data, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T\)
where \(\boldsymbol{W}_{obs}\) and \(\boldsymbol{W}_{mis}\) are the
observed and missing data of the covariate gender, respectively. In this
setting, \(\boldsymbol{W}_{obs}\) is a \(n_{extant} \times 1\) vector
and \(\boldsymbol{W}_{obs}\) is a \(n_{fossil} \times 1\) vector,
perfectly corresponding to the observable gender of the modern extants
and the unobservable gender of the fossil, respectively.

We impute the missing gender for the fossil data by sampling from the
posterior predictive distribution
\(P(\boldsymbol{W}_{mis}|\boldsymbol{W}_{obs}, \boldsymbol{X})\) using
the multiple imputation with chained equations (MICE) algorithm (Buuren
and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) with
predictive mean matching. The imputation algorithm is run to obtain a
total of \(M = 100\) completed datasets.

@@

After imputing sex, we fit an OU model for different phenotypes of
interest. We only look at \(n_{fossil}\) fossil observations.\\
WLOG we choose one of the columns of \(X\) to be the target phenotype
and our analysis is repeated for all phenotypes of interest.

Let \(W_{ti}\) and \(X_{ti}\) be the gender and phenotype of interest
for the \(i\)-th observation in year \(t\) for \(t = 1, \ldots, 18\) and
\(i = 1\ldots,n_{t}\). We assume that

\[
\begin{aligned}
X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\mu_{ft},\sigma^2) & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{mt},\sigma^2) & W_{ti} = \text{Male} \end{array}\right.
\end{aligned}
\] We then fit the time and gender dependent means, \(\mu_{ft}\) and
\(\mu_{mt}\), to an autoregressive process of lag 1, which is the
discretization of an OU process. For \(t = 2,\ldots,T\) and
\(g \in \{f,m\}\), the AR1 model for the mean structure defined as is

\[
\begin{aligned}
\mu_{gt} \overset{iid}{\sim}\mathcal{N}(\kappa \mu_{g,t-1} + (1 - \kappa)\theta_g, \tau^2).
\end{aligned}
\] In this setting, \(\theta_g, g \in \{f,m\}\) is the defined as the
overall long-run mean of the phenotype of interest for each gender, and
\(\kappa\) represents how quickly over time \(\mu_{gt}\) reverts back to
\(\theta_g\); the larger the value of \(\kappa\), the slower
\(\mu_{gt}\) reverts back to \(\theta_g\).

We also set

\[
\begin{aligned}
\mu_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right).
\end{aligned}
\] This model choice is to preserve stationarity,
i.e.~\(p(\mu_{gs}) = p(\mu_{gt})\) for \(s,t \in \{1,\dots,T\}\).

Priors: \[
\sigma \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}}
\] \[
\tau \sim \mathcal{N}(0,100)I_{\{\tau > 0\}}
\]

\[
\kappa \sim \mathcal{N}(0,25)I_{\{0 < \kappa < 1\}} \\
\] \[
\theta_g \sim \mathcal{N}(54,625)I_{\{\theta_g > 0\}} \\
\]

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-MICE}{}}%
Buuren, Stef van, and Karin Groothuis-Oudshoorn. 2011. {``Mice:
Multivariate Imputation by Chained Equations in r.''} \emph{Journal of
Statistical Software} 45 (3): 1--67.
\url{https://doi.org/10.18637/jss.v045.i03}.

\leavevmode\vadjust pre{\hypertarget{ref-Cornault2022}{}}%
Cornuault, Josselin. 2022. {``{Bayesian Analyses of Comparative Data
with the Ornstein--Uhlenbeck Model: Potential Pitfalls}.''}
\emph{Systematic Biology} 71 (6): 1524--40.
\url{https://doi.org/10.1093/sysbio/syac036}.
